Stateless Model Checking with Data-Race Preemption Points

### 2016-03-22 oopsla-16/evaluation.tex
Move discussion of static DR as seed sets to warp zone.

We represented the MC state-of-the-art with 3
%different
configurations of stand-alone \landslide~on the same test suite:
\begin{itemize}
	\item {\bf SSS-MC-DPOR}: Single state space using the maximal PP set from \sect{\ref{sec:initial-pp}}, explored with plain DPOR \cite{dpor},
	\item {\bf SSS-MC-ICB}: With PPs as above, but instead using ICB \cite{chess-icb} with BPOR \cite{bpor} to find bugs faster, and
	\item {\bf SSS-MC-Shared-Mem}: % TODO: better name?
		Using ICB+BPOR, configured to insert PPs on any shared memory access
		(determined at runtime, excluding threads' accesses to their own stacks),
		which in principle includes all possible data-race PPs.
\end{itemize}
Because parallelizing DPOR/ICB during SSS-MC is an open research problem \cite{parallel-dpor},
we instead gave control experiments 10 hours per test with 1 CPU.
%To match this,
\quicksand~reports the CPU-time spent alongside the wall-clock time for a direct comparison
(although, with the growing importance of multicore, \quicksand's inherent parallelism is a convenient benefit).
%Finally,
All tests ran on 12-core 3.2 GHz Xeon W3670 machines. %with 12GB of RAM.

@@

%{\bf Variance.}
% TODO CAMREADY: Probably will need at least a 3rd run of dis.
Because \quicksand~can be nondeterministic in its scheduling of state spaces,
we ran the QS-DR tests a second time to quantify the variance.
We found the area between the two QS-DR curves to be
% TODO
{\bf XX\%}
of the area between QS-DR and SSS-MC-ICB (the next closest),
% TODO cite
which we consider an insignificant amount of variance.
}

@@

% TODO: fix this table.
\begin{table*}[t]
	\begin{center}
		\small
	\begin{tabular}{r|c||c|c|c|c|c||c|c||c|c}
		% TODO: Update SSSMC column to have more favourable ICB numbers.
		& {\bf Num.} & {\bf DPOR} & {\bf QS} & {\bf DR-only} & {\bf Nondet.} & {\bf Malloc-} &
		{\bf Mutual} & {\bf Avg. tested} & {\bf Total} & {\bf Untested} \\
		{\bf Test name} & {\bf tested} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR bugs} & {\bf recycle DRs} &
		{\bf time-outs} & {\bf subset SSes} & {\bf DR PPs} & {\bf DR PPs} \\
		\hline
		% FIXME: update FRMs
				% #test  sssmc   qs      dronly  nondets  FRMs   timeout comp.SSes DRPPs untested DRPPs
		{\tt bcast}	& 79	& 7	& 11	& 4	& 4	& 51	& 0	& -	& 912	& 107	\\
		{\tt join} 	& 79	& 14 	& 24	& 10	& 6	& 333	& 5	& 60.8	& 781	& 292	\\
		{\tt mx} 	& 79	& 1	& 10	& 9	& 1	& 7	& 0	& -	& 829	& 1	\\
		{\tt sem} 	& 79	& 10	& 16	& 6	& 5	& 140	& 29	& 73.3	& 753	& 279	\\
		{\tt signal} 	& 79	& 4	& 10	& 7	& 6	& 180	& 10	& 54.2	& 1118	& 391	\\
		{\tt rwlock} 	& 79	& 26	& 28	& 2	& 2	& 125	& 19	& 29.6	& 915	& 634	\\
		\hline
		{\tt sched} 	& 59	& 1	& 7	& 6	& 4	& 0	& 0	& -	& 144	& 3	\\
		{\tt alarm} 	& 44	& 5	& 21	& 17	& 3	& 35 	& 22	& 9.1	& 115	& 89	\\
		{\tt wait} 	& 52	& 23	& 30	& 7	& 1	& 31	& 8	& 28.0	& 142	& 31	\\
		\hline
		{\bf Total}	& 629	& 91	& 157	& 68	& 32	& 902	& 93	& 42.6 	& 5709	& 1827	\\
	\end{tabular}
	\end{center}
	\caption{Summary of bugs and data races found by each test program.
		% TODO
		DPOR is the control; QS is \quicksand.
		``DR-only bugs'' counts among \quicksand's bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
	among those, ``Nondeterministic DR bugs'' counts how many candidates required MC integration to identify (\sect{\ref{sec:eval-dr}}).
	``Malloc-recycle DRs'' counts how many false positives we suppressed (\sect{\ref{sec:eval-dr}}).
	``Mutual time-outs'' counts how often both SSS-MC and \quicksand~timed out with no bug found;
	among those, ``Average tested subset SSes'' counts how many partial verifications \quicksand~provided on average for each test (\sect{\ref{sec:eval-sssmc}}).
	``Total DR PPs'' counts how many unique data-racing instructions we identified among tests where we found no bugs;
	among those, ``Untested DR PPs'' counts how many could not be checked in the time limit (\sect{\ref{sec:future}}).
		}
	\label{tab:drbugs}
\end{table*}

@@

\revision{
On the other hand,
comparing QS-DR to SSS-MC-Shared-Mem shows that Iterative Deepening thoroughly outperforms ICB when shared-memory preemptions come into play.
%We attribute this to the fact that
Statically configuring a PP for every shared memory access in advance
produces orders of magnitude more PPs than
waiting for an access to be identified as part of a (potential) data race at runtime.
%
In principle, DPOR and BPOR should identify and prune any equivalences arising from extraneous PPs on non-conflicting accesses.
However, in practice,
%we found that
the sheer number of accesses during each new execution (often thousands) added significant performance overhead to the MC when computing DPOR and backtracking\footnote{
	\revision{The bug-finding performance of SSS-MC-ICB approach could be improved by heuristically
	%preempting only on data-race candidates found in advance by a single-pass analysis
	using a single-pass analysis to find data-race PP candidates in advance
	\cite{portend}, but this sacrifices soundness, as shown in \sect{\ref{sec:eval-dr}}.}
}.
Iterative Deepening avoids this overhead by waiting until runtime to identify fewer, more relevant PPs dynamically,
and is hence more suitable for MC with data-race PPs.}

>>>>

We represented the MC state-of-the-art with 3
%different
configurations of stand-alone \landslide~on the same test suite:
\begin{itemize}
	\item {\bf SSS-MC-DPOR}: Single state space using the maximal PP set from \sect{\ref{sec:initial-pp}}, explored with plain DPOR \cite{dpor},
	\item {\bf SSS-MC-ICB}: With PPs as above, but instead using ICB \cite{chess-icb} with BPOR \cite{bpor} to find bugs faster, and
	\item {\bf SSS-MC-Shared-Mem}: % TODO: better name?
		Using ICB+BPOR, configured to insert PPs on any shared memory access
		(determined at runtime, excluding threads' accesses to their own stacks),
		which in principle includes all possible data-race PPs.
\end{itemize}
Because parallelizing DPOR/ICB during SSS-MC is an open research problem \cite{parallel-dpor},
we instead gave control experiments 10 hours per test with 1 CPU.
%To match this,
\quicksand~reports the CPU-time spent alongside the wall-clock time for a direct comparison
(although, with the growing importance of multicore, \quicksand's inherent parallelism is a convenient benefit).
%Finally,
All tests ran on 12-core 3.2 GHz Xeon W3670 machines. %with 12GB of RAM.
}

@@

\revision{
%{\bf Variance.}
% TODO CAMREADY: Probably will need at least a 3rd run of dis.
Because \quicksand~can be nondeterministic in its scheduling of state spaces,
we ran the QS-DR tests a second time to quantify the variance.
We found the area between the two QS-DR curves to be
% TODO
{\bf XX\%}
of the area between QS-DR and SSS-MC-ICB (the next closest),
% TODO cite
which we consider an insignificant amount of variance.
}

@@

% TODO: fix this table.
\begin{table*}[t]
	\begin{center}
		\small
	\begin{tabular}{r|c||c|c|c|c|c|c||c|c||c|c}
		% TODO: Update SSSMC column to have more favourable ICB numbers.
		& {\bf Num.} & {\bf DPOR} & {\bf ICB} & {\bf QS} & {\bf DR} & {\bf Nondet.} & {\bf Malloc} & {\bf Mutual} & {\bf Avg. tested} & {\bf Total} & {\bf Untested} \\
		{\bf Test} & {\bf tested} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf bugs} & {\bf DR bugs} & {\bf DRs} & {\bf timeouts} & {\bf subset SSes} & {\bf DR PPs} & {\bf DR PPs} \\
		\hline
		% FIXME CAMREADY: update FRMs
					% TODO: update to be ShMem not DPOR
											% TODO: recount with ICB
				% #test  sssmc   icb      qs      dronly  nondets  FRMs   timeout comp.SSes DRPPs untested DRPPs
		{\tt bcast}	& 79	& 7	& TODO	& 11	& 4	& 4	& 51	& 0	& -	& 912	& 107	\\
		{\tt join} 	& 79	& 14 	&	& 24	& 10	& 6	& 333	& 5	& 60.8	& 781	& 292	\\
		{\tt mx} 	& 79	& 1	&	& 10	& 9	& 1	& 7	& 0	& -	& 829	& 1	\\
		{\tt sem} 	& 79	& 10	&	& 16	& 6	& 5	& 140	& 29	& 73.3	& 753	& 279	\\
		{\tt signal} 	& 79	& 4	&	& 10	& 7	& 6	& 180	& 10	& 54.2	& 1118	& 391	\\
		{\tt rwlock} 	& 79	& 26	&	& 28	& 2	& 2	& 125	& 19	& 29.6	& 915	& 634	\\
		\hline
		{\tt sched} 	& 59	& 1	&	& 7	& 6	& 4	& 0	& 0	& -	& 144	& 3	\\
		{\tt alarm} 	& 44	& 5	&	& 21	& 17	& 3	& 35 	& 22	& 9.1	& 115	& 89	\\
		{\tt wait} 	& 52	& 23	&	& 30	& 7	& 1	& 31	& 8	& 28.0	& 142	& 31	\\
		\hline
		{\bf Total}	& 629	& 91	&	& 157	& 68	& 32	& 902	& 93	& 42.6 	& 5709	& 1827	\\
	\end{tabular}
	\end{center}
	\caption{Summary of bugs and data races found by each test program.
		% TODO
		DPOR is the control; QS is \quicksand.
		``DR bugs'' counts among \quicksand's bugs how many required data-race PPs to expose (\sect{\ref{sec:eval-sssmc}});
	among those, ``Nondeterministic DR bugs'' counts how many candidates required MC integration to identify (\sect{\ref{sec:eval-dr}}).
	``Malloc DRs'' counts how many malloc-recycle false positives we suppressed (\sect{\ref{sec:eval-dr}}).
	``Mutual timeouts'' counts how often both SSS-MC and \quicksand~timed out with no bug found;
	among those, ``Average tested subset SSes'' counts how many partial verifications \quicksand~provided on average for each test (\sect{\ref{sec:eval-sssmc}}).
	``Total DR PPs'' counts how many unique data-racing instructions we identified among tests where we found no bugs;
	among those, ``Untested DR PPs'' counts how many could not be checked in the time limit (\sect{\ref{sec:future}}).
		}
	\label{tab:drbugs}
\end{table*}

@@

\revision{
On the other hand,
comparing QS-DR to SSS-MC-Shared-Mem shows that Iterative Deepening thoroughly outperforms ICB when shared-memory preemptions come into play.
%We attribute this to the fact that
Statically configuring a PP for every shared memory access in advance
produces orders of magnitude more PPs than
waiting for an access to be identified as part of a (potential) data race at runtime.
%
In principle, DPOR and BPOR should identify and prune any equivalences arising from extraneous PPs on non-conflicting accesses.
However, in practice,
%we found that
the sheer number of accesses during each new execution (often thousands) added significant performance overhead to the MC when computing DPOR and backtracking.
Iterative Deepening avoids this overhead by waiting until runtime to identify fewer, more relevant PPs dynamically,
and is hence more suitable for MC with data-race PPs.}

### 2016-03-22 oopsla-16/related.tex
Move discussion of static DR as seed sets to warp zone.

{\bf \revision{Reduction}~techniques.} Various improvements or alternatives to DPOR have been developed, such as Dynamic Interface Reduction \cite{demeter}, Maximal Causality Reduction \cite{mcr},
%DPOR for relaxed memory models \cite{tsopso},
SAT-directed MC \cite{satcheck},
\revision{and R4 for event-driven applications \cite{r4}}.
These are all compatible with our technique.
Recent work \cite{tsopso} has extended DPOR for relaxed memory models \cite{memory-consistency-models},
which we do not yet account for in our proofs (\sect{\ref{sec:soundness}}).
Parrot \cite{parrot} combines MC with a partially-determinizing runtime for further reduction, but still, fewer than half the non-trivial \revision{tests}~in their evaluation could be completed,
\revision{which motivates \quicksand's CPU-budget-oriented approach.}
%providing a strong argument for \quicksand.

\revision{{\bf Restricting preemptions.}
Preemption Sealing \cite{sealing} pre\-sents a mechanism similar to the {\tt within\_function} command (\sect{\ref{sec:landslide}}) for users to manually restrict preemptions.
It demonstrates the need to consider subsets of PPs,
as well as developers' willingness to limit a test case's scope so the resulting state space may be fully verified (\sect{\ref{sec:totalverif}}).
%It justifies our assumption that expert
%programmers can customize small test cases which, when fully verified,
%provide a correctness assurance much stronger than the assurance from
%a partially-complete exploration of a huge state space.
Probabilistic Concurrency Testing (PCT) \cite{randomized-scheduler}
is a randomized algorithm that can quantify the probability of uncovering bugs.
PCT is optimized for tests with impossibly large state spaces,
while for more approachable tests,
its incompatibility with DPOR would significantly delay completion.
Future work could use ETAs to heuristically switch between DPOR and PCT}.
Finally, Iterative Context Bounding (ICB) \cite{chess-icb} is most similar to our work,
as both approaches provide a partial verification
%on some subset of interleavings
when full completion is intractable (\sect{\ref{sec:future}}).
However, ICB is limited to a fixed set of PPs,
\revision{cannot estimate remaining time to total verification},
and to our knowledge no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.

>>>>

% TODO: Expand on MCR and SATcheck.
{\bf \revision{Reduction}~techniques.} Various improvements or alternatives to DPOR have been developed, such as Dynamic Interface Reduction \cite{demeter}, Maximal Causality Reduction \cite{mcr},
%DPOR for relaxed memory models \cite{tsopso},
SAT-directed MC \cite{satcheck},
\revision{and R4 for event-driven applications \cite{r4}}.
These are all compatible with our technique.
Recent work \cite{tsopso} has extended DPOR for relaxed memory models \cite{memory-consistency-models},
which we do not yet account for in our proofs (\sect{\ref{sec:soundness}}).
Parrot \cite{parrot} combines MC with a partially-determinizing runtime for further reduction, but still, fewer than half the non-trivial \revision{tests}~in their evaluation could be completed,
\revision{which motivates \quicksand's CPU-budget-oriented approach.}
%providing a strong argument for \quicksand.

\revision{{\bf Restricting preemptions.}
Preemption Sealing \cite{sealing} pre\-sents a mechanism similar to the {\tt within\_function} command (\sect{\ref{sec:landslide}}) for users to manually restrict preemptions.
It demonstrates the need to consider subsets of PPs,
as well as developers' willingness to limit a test case's scope so the resulting state space may be fully verified (\sect{\ref{sec:totalverif}}).
%It justifies our assumption that expert
%programmers can customize small test cases which, when fully verified,
%provide a correctness assurance much stronger than the assurance from
%a partially-complete exploration of a huge state space.
Probabilistic Concurrency Testing (PCT) \cite{randomized-scheduler}
is a randomized algorithm that can quantify the probability of uncovering bugs.
PCT is optimized for tests with impossibly large state spaces,
while for more approachable tests,
its incompatibility with DPOR would significantly delay completion.
Future work could use ETAs to heuristically switch between DPOR and PCT}.
Finally, Iterative Context Bounding (ICB) \cite{chess-icb} is most similar to our work,
as both approaches provide a partial verification
%on some subset of interleavings
when full completion is intractable (\sect{\ref{sec:future}}).
However, ICB is limited to a fixed set of PPs,
\revision{cannot estimate remaining time to total verification,
and cannot incorporate data-race PPs not statically coded in advance}.
%and to our knowledge no algorithm has been proposed to dynamically add data-race PPs during a test with ICB.

### 2016-03-22 oopsla-16/warpzone.tex
Move discussion of static DR as seed sets to warp zone.

>>>>

\revision{
{\bf Integration with static data-race analysis.}
In \sect{\ref{sec:eval-sssmc}}, we evaluated SSS-MC's ability to find data-race-induced failures
by configuring a static predicate to preempt on any non-stack memory access.
This introduced up to thousands of new PPs on each new test execution, with a prohibitive performance impact.
While this performance could be improved by
%relaxing the preemption strategy,
instead using a static or single-pass analysis to find data-race candidate PPs in advance \cite{portend},
this strategy sacrifices the soundness of the verification guarantee, as shown in \sect{\ref{sec:eval-dr}}.
However, \quicksand~itself could employ static data-race analysis \cite{racerx} in future work.
Statically-identified data race candidates could heuristically be included in our ``seed'' PP sets (\sect{\ref{sec:initial-pp}}),
enabling \quicksand~to focus on the most suspicious races immediately, rather than waiting for them to be identified after potentially many iterations of MC.
}
